Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-27 paper/paper.tex
Minor

In the Coq mechanized proof, the definition of $[[Interp I i A S]]$ has type
\texttt{Prop}, when \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S}
has type \texttt {tm -> Prop}.  However, if desired, we could consistently
replace the use of \texttt{Prop} with Coq's predicative sort
\texttt{Type}. This variation of the definition could be used to define the
interpretation for any \emph{finite} number of universes. The use of
\texttt{Type} becomes troublesome only when we attempt to define
$[[InterpR i A S]]$, the top-level logical relation defined in
Figure~\ref{fig:logrelrec} that recursively calls itself at smaller universe
levels. Therefore, the one feature of \lang{} that truly requires impredicativity
is its countable universe hierarchy. \scw{I'm not sure the last part of this sentence
adds much: ``Although, lemmas such as the
admissibility of \rref{I-PiAlt} (Lemma~\ref{lemma:piintroalt}) not only simply
our mechanization, but also give better intuition on what our logical relation
means.'' Yes, we need impredicativity to define these lemmas, but without impredicativity
our logical relation means something else.}

@@

Finally, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$
are equal, it suffices to show the extensional property that $\forall
x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
occasionally in our presented proofs.
In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm ->
Prop}, a predicate over \lang{} terms.
In axiom-free Coq, predicates do not come with the extensionality
property. Given two predicates $P$ and $Q$, we can not conclude that
$P = Q$ when given a proof of $\forall
x, P(x) \iff Q(x)$. While
predicate extensionality is not necessary in our development, we
assume it as an axiom since it is known to be consistent with Coq's
metatheory and helps bridge the gap between our mechanization and the
proofs we present.

@@

The soundness of the algorithm is immediate. The completeness of the
algorithm is justified the confluence property of the untyped
reduction relation. Suppose $[[a <=> b]]$, $[[a =>+ f0]]$, $[[b =>+
f1]]$ but $[[f0]]$ is syntactically distinct from $[[f1]]$. By the transitivity of convertibility, we have $[[f0 <=> f1]]$.
Because $[[f0]]$ and $[[f1]]$ are both in
$\beta\eta$-normal forms and can only reduce to themselves, we must
have $[[f0 = f1]]$. Contradiction.

@@

However, with a little more effort, it is possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal form and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data}.

@@

\citet{decagda} mechanizes in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$-law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidirectional type system.

\citet{anand2014towards} mechanizes the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
extensional type theory with features such as dependent functions,
inductive types, and a full universe hierarchy. \citet{nbeincoq}
mechanizes the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the
only base type from the object language, large elimination is not
supported despite the one predicative universe.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further shows it is possible
to encode a finite universe hierarchy without the use of
either impredicativity or induction-recursion.
% \yl{commented out because I think this paragraph is too subjective}
% However, instead of explaining the inductive
% logical relation as a convoluted workaround to the lack of
% induction-recursion in Coq, we give a self-contained explanation of
% our logical relation.

>>>>

In the Coq mechanized proof, the definition of $[[Interp I i A S]]$ has type
\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S}
has type \texttt {tm -> Prop}.  However, if desired, we could consistently
replace the use of \texttt{Prop} with Coq's predicative sort
\texttt{Type}. This variation of the definition could be used to define the
interpretation for any \emph{finite} number of universes. The use of
\texttt{Type} becomes troublesome only when we attempt to define
$[[InterpR i A S]]$, the top-level logical relation defined in
Figure~\ref{fig:logrelrec} that recursively calls itself at smaller universe
levels. Therefore, the one feature of \lang{} that truly requires impredicativity
is its countable universe hierarchy. \scw{I'm not sure the last part of this sentence
adds much: ``Although, lemmas such as the
admissibility of \rref{I-PiAlt} (Lemma~\ref{lemma:piintroalt}) not only simply
our mechanization, but also give better intuition on what our logical relation
means.'' Yes, we need impredicativity to define these lemmas, but without impredicativity
our logical relation means something else.}

@@

Finally, in set theory, to show that two sets $[[S0]]$ and $[[S1]]$
are equal, it suffices to show the extensional property that $\forall
x, x \in [[S0]] \iff x \in [[S1]]$. We leverage this fact
occasionally in our presented proofs.
In Coq, sets of terms ($[[PowerSet STm]]$) is encoded as the type \texttt{tm ->
Prop}, a predicate over \lang{} terms.
In axiom-free Coq, predicates do not come with the extensionality
property. Given two predicates $P$ and $Q$, we cannot conclude that
$P = Q$ when given a proof of $\forall
x, P(x) \iff Q(x)$. While
predicate extensionality is not necessary in our development, we
assume it as an axiom since it is known to be consistent with Coq's
metatheory and helps bridge the gap between our mechanization and the
proofs we present.

@@

The soundness of the algorithm is immediate. The completeness of the
algorithm is justified the confluence property of the untyped
reduction relation. Suppose $[[a <=> b]]$, $[[a =>+ f0]]$, $[[b =>+
f1]]$ but $[[f0]]$ is syntactically distinct from $[[f1]]$. By the transitivity of convertibility, we have $[[f0 <=> f1]]$.
Because $[[f0]]$ and $[[f1]]$ are both in
$\beta\eta$-normal forms and can only reduce to themselves, we must
have $[[f0 = f1]]$, which is a contradiction.

@@

However, % with a little more effort,
it is possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal forms and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data.}.

@@

\citet{decagda} mechanize in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and typed judgmental equality with function
$\eta$ law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq.
\citet{martin-lof-a-la-coq} manages to encode the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidirectional type system.

\citet{anand2014towards} mechanize the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The metatheory is an
extensional type theory with features such as dependent functions,
inductive types, and a full universe hierarchy. \citet{nbeincoq}
mechanize the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the
only base type from the object language, large elimination is not
supported despite the one predicative universe.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further shows it is possible
to encode a finite universe hierarchy without the use of
either impredicativity or induction-recursion.
% \yl{commented out because I think this paragraph is too subjective}
% However, instead of explaining the inductive
% logical relation as a convoluted workaround to the lack of
% induction-recursion in Coq, we give a self-contained explanation of
% our logical relation.
